---
title: "Problem Set 3"
author: "Ben Funk"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(tidyverse)
library(HistData)
```

## Introduction

These questions were rendered in R markdown through RStudio (<https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf>, <http://rmarkdown.rstudio.com> ).

Please generate your solutions in R markdown and upload both a knitted doc, docx, or pdf document in addition to the Rmd file.

The questions in this problem set use material from the slides on discrete probability spaces and the Rmd "Discrete_Probability_Distributions_2_3_3.Rmd".

### Collaboration

(5 points)

Other students consulted on assignment. Please write none if you worked by yourself: None

### AI

(5 points)

AI tools used in this assignment. Please write none if you did not use any AI tools: GPT o3

## Introduction

In this problem set, we will look at another approach to the hypothesis testing situation from problem set 2. To do this, we will use the idea of *conditional probability*. In general, given a probability space $(S,M,P)$ and an event $C$, we can define a related probability space $(S_C,M_C,P_C)$ by taking $S_C=S$,$M_C=M$. To define $P_C$, for any $A\in M_C$ set $P_C(A)= P(A\cap C)/P(C)$. This is also called the conditional probability of $A$ occurring given that $C$ has occurred, and is written $P(A|C)$.

The study

Kannan S, Bruch JD, Song Z. Changes in Hospital Adverse Events and Patient Outcomes Associated With Private Equity Acquisition. JAMA. 2023;330(24):2365--2375. <doi:10.1001/jama.2023.23147>

found that the rate of hospital-acquired conditions (HAC) at private equity hospitals increased from 18.1 per 10,000 eligible hospitalizations before acquisition to 22.1 per 10,000 eligible hospitalizations after acquisition.

This problem set, as did problem set 2, will use those rates to illustrate an application of binomial probability spaces in a much simpler setting than addressed in the study.

### Frame Question

For the purposes of the questions on this problem set, we assume outcomes of eligible hospitalizations are independent and the rate of hospital-acquired conditions has some value $p$ that does not depend on whether the observation was made before or after private equity-acquisition. The binary outcome of whether a hospital-acquired condition occurred is noted for each eligible hospitalization.

Suppose that there were $n_b=3,000$ eligible hospitalizations before acquisition and $n_a=2,000$ eligible hospitalizations after acquisition.(These numbers are reduced from problem set 2 to speed and simplify the numeric estimates.)

Explicitly, the null model $(S,M,P)$ is defined by setting $S=\left\{(a,b): a\in \{ 0,...2000\} \wedge b\in \{ 0,...3000\}\right\}$, the density $f((a,b))=dbinom(a,2000,p)dbinom(b,3000,p)$, using R notation, and $M$ is the power set of $S$. This represents the experiment of sampling $Binomial(2000,p)$ to get $a$ and $Binomial(3000,p)$ to get $b$ and recording $(a,b)$ as an outcome. Note that the probability that the total number of HACs, $a+b$, equals some $c$ is given by the binomial distribution with size $5000$ and probability $p$ because it is the number of occurrences of HACs in $5000$ independent trials in which the probability of an occurrence is $p$.

### Question 1, part 1

(5 points)

To begin to define a hypothesis test directly from this model, let $A$ be the event $\{(x,y)\in S:x=a\}$ and let $C$ be the event $\{(x,y)\in S:x+y=c\}$. Please calculate the probability of $A$ given $C$ using this null model. Please include the derivation.

Please give your answer in terms of $a$, $c$, $n_a$, and $n_b$. You will note that $p$ does not appear in the formula.

#### your answer here

Conditional probability is given as: $P(A \mid C) = \frac{P(A \cap C)}{P(C)}.$

This means from the null model, the outcome of a given c and a occurs can be found through: $$P(A \cap C) = \binom{n_a}{a} p^a (1-p)^{n_a - a} \cdot \binom{n_b}{c-a} p^{c-a} (1-p)^{n_b - (c-a)}= \binom{n_a}{a}\binom{n_b}{c-a} p^c (1-p)^{n_a+n_b-c}.$$

Because c successes occur we also know $$P(C) = \binom{n_a+n_b}{c} p^c (1-p)^{n_a+n_b-c}.$$

meaning that all of this simplifies down to the formula: $$P(A \mid C) = \frac{\binom{n_a}{a}\binom{n_b}{c-a}}{\binom{n_a+n_b}{c}}.$$

### Q1, part 2

(10 points)

Consider the two functions below intended to calculate approximate numerical values for probability of $A$ given $C$ under the null model.

```{r}

n.after<-2000 # number of eligible hospitalizations in after acquisition
n.before<-3000 # number of eligible hospitalizations in before acquisition
tot<-9


prob.a.given.c<-function(a,c,n.a,n.b){
  b<-c-a
  return(exp(log(choose(n.a,a))+log(choose(n.b,b))-log(choose(n.a+n.b,c))))
  
}

prob.a.given.c.est<-function(a,c,n.a,n.b){
  return(dbinom(a,c,prob=n.a/(n.a+n.b)))
  
}

a_values <- 0:tot
direct_results <- sapply(a_values, function(a) prob.a.given.c(a, tot, n.after, n.before))
est_results    <- sapply(a_values, function(a) prob.a.given.c.est(a, tot, n.after, n.before))

comparison_df <- data.frame(a = a_values,
                            Direct = direct_results,
                            Estimation = est_results)
print(comparison_df)

are_identical <- all.equal(direct_results, est_results)
are_identical
```

Do the functions give the same results for all values of $a$, $c$, $n_a$, and $n_b$? If so, why? If not, please give values of $a$, $c$, $n_a$, and $n_b$ for which the functions give different results.

#### your answer here

They are close but there is still some variation in each of the answers., but for each value of a they are extremely similar, and an average difference of

```         
0.00085
```

## Question 1, part 3

(10 points)

We will continue the construction of a hypothesis test defined by the null model above using the function `prob.a.given.c` to give a numerical value for the probability of $A$ given $C$ under the null model.

For an outcome $(a,b)$, we will reject the null hypothesis at the 0.05 level, say, if the probability of $\{(x,c-x):x\leq a\}$, given $C=\{(x,y)\in S:x+y=c\}$, is less than or equal to 0.025. We will also reject the null at the 0.05 level if the probability of $\{(x,c-x):x\geq a\}$ given $C$ is less than or equal to 0.025.

```{r}
n.a<-n.after
n.b<-n.before
c<-10
# Function to calculate the probability of A given C for a in 0 through the value of c defining the event C
p.true.c<-function(c,n.a,n.b){
  return(prob.a.given.c(0:c,c,n.a,n.b))
}

```

The function `p.reject` calculates the probability of an outcome $(a,b)$ given $c$ that would lead to rejection of the null hypothesis at the 0.05 level. The variable `p.reject.vec` applies this function to the values of $c$ from 1 to 150.

```{r}

p.reject<-function(c,n.a,n.b){
  q<-n.a/(n.a+n.b)
  # calculate the probability of a given c for a in 0 through c
  p.true<-p.true.c(c,n.a,n.b)
  # calculate the cumulative sum of the probabilities
  p.true.cum<-cumsum(p.true)
  # Find the largest a for which the of a value of a or less given c is less than .025
  if(p.true[1]<.025){
    lower.tail<-max(which(p.true.cum<.025)) # the largest index of the cumulative sum with value less than .025
    lower.sum<-sum(p.true[1:lower.tail]) # the sum of the probabilities of a value of a or less
  } else {
    lower.sum<-0
  }
  p.true.rev<-rev(p.true) # reverse the order of the probabilities
  p.true.rev.cum<-cumsum(p.true.rev)
  if(p.true.rev[1]<.025){
    upper.tail<-max(which(p.true.rev.cum<.025))
    upper.sum<-sum(p.true.rev[1:upper.tail])
  } else {
    upper.sum<-0
  }
  
  return(lower.sum+upper.sum)
}

p.reject(75,n.after,n.before)

p.reject.vec<-sapply(1:150,p.reject,n.a=n.after,n.b=n.before)
p.reject.vec[10]

```

What is the type 1 error rate of this test given c=10?

#### your answer here

it is

```         
0.01823281
```

### Question 1, part 4

(10 points)

The vector `c.vec` contains the probabilities of $c=1$ through $c=150$ given the null model and $p=0.00201$.

```{r}
HAC.rate<-(18.1/10000+22.1/10000)/2
c.vec<-dbinom(1:150,n.a+n.b,HAC.rate)

(c.vec*p.reject.vec)[1:15]
```

With this value of $p$, what is the probability of rejecting the event that both $c=10$ and the null hypothesis is rejected at the 0.05 level? That is, what is the probability of the event $\{(a,b):a+b=10 \text{ and } a \text{ results in rejection of the null at the 0.05 level} \}$?

#### your answer here

```{r}
(c.vec * p.reject.vec)[10]
```

### Question 1, part 5

(5 points)

Note that the probability of the event $D=\{(x,y):x+y>150\}$ extremely small.

```{r}
pbinom(150,n.a+n.b,prob=HAC.rate,lower.tail=FALSE)

```

Please give a numerical estimate for the probability of a type 1 error under the null model using this method and this value of $p$ based on the calculations above.

#### your answer here

```{r}
type1_error_partial <- sum(c.vec * p.reject.vec)

tail_prob <- pbinom(150, n.a + n.b, HAC.rate, lower.tail=FALSE)

type1_error_est <- type1_error_partial + tail_prob

type1_error_est
```
